---
title: "Quant 2, Lab 11"
subtitle: "Experiments and DeclareDesign"
author: "Sylvan Zheng"
output: html_document
---

# DeclareDesign

Material adapted from https://book.declaredesign.org/declaration-diagnosis-redesign/diagnosing-designs.html

```{r}
pacman::p_load(DeclareDesign, tidyverse, future, future.apply)
theme_update(text = element_text(size = 24))
```

## Simple Random Experiment

```{r}
# Declare a simple potential outcomes model
N <- 100
tau <- 0.2

declaration_18.1 <-
    declare_model(
        N = N,
        U = rnorm(N),
        potential_outcomes(Y ~ tau * Z + U)
    ) +
    declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
    declare_assignment(Z = complete_ra(N, prob = 0.5)) +
    declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
    declare_estimator(Y ~ Z, inquiry = "ATE")

dd <- diagnose_design(declaration_18.1)
dd
dd$simulations_df %>% head()
hist(dd$simulations_df$estimate)
hist(dd$simulations_df$p.value)
# Power
sum(dd$simulations_df$p.value <= 0.05) / 500
```

## `redesign`: Evaluating Multiple Designs

### Vary the sample size

```{r}
dd.n <- declaration_18.1 %>%
    redesign(
        N = seq(100, 1000, 100)
    ) %>%
    diagnose_designs()

dd.n$diagnosands_df %>% select(N, sd_estimate, power)

# plot the average ATE estimate and 95% CIs
dd.n$diagnosands_df %>% ggplot(aes(x = N, y = mean_estimate, ymin = mean_estimate - 1.96 * sd_estimate, ymax = mean_estimate + 1.96 * sd_estimate)) +
    geom_point(size = 4) +
    geom_errorbar(width = 0, linewidth = 1)

# plot the power curve
dd.n$diagnosands_df %>% ggplot(aes(x = N, y = power)) +
    geom_line(linewidth = 2)
```

### Vary the effect size

```{r}
dd.tau <- declaration_18.1 %>%
    redesign(
        N = seq(100, 1000, 100),
        tau = c(0.1, 0.2, 0.3)
    ) %>%
    diagnose_designs()

dd.tau$diagnosands_df %>% ggplot(aes(x = N, y = power, color = as.factor(tau))) +
    geom_line(linewidth = 2) +
    labs(color = "tau")
```

## Covariate Adjustment

```{r}
N <- 100
r_sq <- 0

declaration_18.2 <-
    declare_model(
        N = N,
        draw_multivariate(c(U, X) ~ MASS::mvrnorm(
            n = N,
            mu = c(0, 0),
            Sigma = matrix(c(1, sqrt(r_sq), sqrt(r_sq), 1), 2, 2)
        )),
        potential_outcomes(Y ~ 0.1 * Z + U)
    ) +
    declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
    declare_assignment(Z = complete_ra(N)) +
    declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
    declare_estimator(
        Y ~ Z,
        covariates = ~X, .method = lm_lin, inquiry = "ATE"
    )

# Vary the rsq value along with N
rdd <- declaration_18.2 %>%
    redesign(
        r_sq = seq(0, 0.8, 0.2),
        N = seq(100, 1000, 100)
    ) %>%
    diagnose_designs(bootstrap_sims = 0)

rdd$duration
rdd$diagnosands_df %>% ggplot(aes(x = N, y = power, color = as.factor(r_sq))) +
    geom_line(linewidth = 2) +
    labs(color = "Rsq")
```

### Comparing Lin estimator to OLS

It can be less efficient to use OLS with covariates in a very specific circumstance:
when the relationship between the covariate and the control and treatment outcomes 
are different AND when the experiment has an "imbalance" in terms of how many 
units are assigned to control

```{r}
prob <- 0.5
# In this example, the control potential outcomes are not related to covariate
# but the treatment potential outcomes are related.
control_slope <- 0
treatment_slope <- 1

declaration_18.3 <-
    declare_model(
        N = N,
        X = runif(N, 0, 1),
        U = rnorm(N, sd = 0.1),
        Y_Z_1 = treatment_slope * X + U,
        Y_Z_0 = control_slope * X + U
    ) +
    declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
    declare_assignment(Z = complete_ra(N = N, prob = prob)) +
    declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
    declare_estimator(Y ~ Z, inquiry = "ATE", label = "DIM") +
    declare_estimator(Y ~ Z + X, .method = lm_robust, inquiry = "ATE", label = "OLS") +
    declare_estimator(Y ~ Z, covariates = ~X, .method = lm_lin, inquiry = "ATE", label = "Lin")

dd.lin <- declaration_18.3 %>%
    redesign(
        prob = seq(0.1, 0.5, 0.1),
        control_slope = c(-1, 0, 1)
    ) %>%
    diagnose_designs()

dd.lin$diagnosands_df %>% ggplot(aes(x = prob, y = sd_estimate, color = estimator)) +
    facet_wrap(~control_slope) +
    geom_line(linewidth = 2)
```

## Block Randomization

```{r}
tau <- 0.2
coef <- 2

block.m <-
    declare_model(
        N = N,
        gender = sample(size = N, c(0, 1), replace = T),
        U = rnorm(N, sd = 1),
        potential_outcomes(Y ~ tau * Z + coef * gender + U)
    ) +
    declare_assignment(
        Z = block_ra(blocks = gender),
    ) +
    declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
    declare_estimator(Y ~ Z, label = "DIM") +
    declare_estimator(
        Y ~ Z,
        covariates = ~gender,
        .method = lm_lin,
        label = "Lin"
    )
simple.m <-
    declare_model(
        N = N,
        gender = sample(size = N, c(0, 1), replace = T),
        U = rnorm(N, sd = 1),
        potential_outcomes(Y ~ tau * Z + coef * gender + U)
    ) +
    declare_assignment(
        Z = simple_ra(N = N, prob = 0.5),
    ) +
    declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
    declare_estimator(Y ~ Z, label = "DIM") +
    declare_estimator(
        Y ~ Z,
        covariates = ~gender,
        .method = lm_lin,
        label = "Lin"
    )
dd.block <- block.m %>%
    redesign(
        N = seq(40, 200, 20),
        coef = c(0.5, 1, 2, 4)
    ) %>%
    diagnose_designs()
dd.simple <- simple.m %>%
    redesign(
        N = seq(40, 200, 20),
        coef = c(0.5, 1, 2, 4)
    ) %>%
    diagnose_designs()
rbind(
    dd.block$diagnosands_df %>% mutate(method = "block"),
    dd.simple$diagnosands_df %>% mutate(method = "simple")
) %>% ggplot(aes(x = N, y = power, color = method)) +
    facet_wrap(~ estimator + coef) +
    geom_line()
```


