---
title: "Quant 2, Week 9"
subtitle: "Panel Imputation Methods"
author: "Sylvan Zheng"
---

```{r}
pacman::p_load(fect, panelView, tidyverse)
devtools::install_github("ebenmichael/augsynth")
devtools::install_github("chadhazlett/kbal")
devtools::install_github("xuyiqing/tjbal")
library(augsynth)
library(tjbal)
data(fect)
# Turnout dataset
# EDR = election-day registration policy
head(turnout)
panelview(turnout, index = c("abb", "year"), Y = "turnout", D = "policy_edr")
panelview(simdata1, index = c("id", "time"), Y = "Y", D = "D")
```

## TWFE Imputation (fe)

- Borusyak, Jaravel & Spiess (2024)
- Model is fit only to pre-treatment and control data and then
used to impute Y0s.

### Implementation by hand with fixest

```{r}
library(fixest)
modl <- feols(Y ~ X1 + X2 | id + time, data = simdata1 %>% filter(D == 0))
simdata1$y0.impute <- predict(modl, newdata = simdata1)
treated <- simdata1 %>% filter(D == 1)
att <- mean(treated$Y) - mean(treated$y0.impute)
att
```

### Implementation using fect package

```{r}
fe <- fect(
    Y ~ D,
    index = c("id", "time"),
    data = simdata1,
    method = "fe",
    se = T,
)
plot(fe)
print(fe)

# With covariates
fe.cov <- fect(
    Y ~ D + X1 + X2,
    index = c("id", "time"),
    data = simdata1,
    method = "fe",
    se = T,
)
plot(fe.cov)
print(fe.cov)
```

## Interactive FE

- Basic idea: instead of simple two way fixed effects to model Y0, do something more sophisticated.
- Choose some low value r (usually r < 5)
- Model r latent factors 
- Say we have 5 units, 3 periods, and 2 factors
- TWFE: Y0 ~ i1 + i2 + i3 + i4 + i5 + t1 + t2 + t3
- IFE: Y0 ~ (r1 + r2)*(i1 + i3 + i4 + i5) + (r1 + r2)*(t1 + t2 + t3)

```{r}
ife <- fect(
    Y ~ D + X1 + X2,
    index = c("id", "time"),
    data = simdata1,
    method = "ife",
    CV = TRUE,
    r = c(0, 5),
    se = T,
)
plot(ife)
print(ife)
```

### Matrix Completion

```{r}
mc <- fect(
    Y ~ D + X1 + X2,
    index = c("id", "time"),
    data = simdata1,
    method = "mc",
    CV = TRUE,
    se = T,
)
plot(mc)
print(mc)
```

### Generalized Synthetic Control (gsynth)

```{r}
gsynth <- fect(
    Y ~ D + X1 + X2,
    index = c("id", "time"),
    data = simdata1,
    method = "gsynth",
    se = T,
    r = c(0, 5),
    CV = T
)
plot(gsynth)
print(gsynth)
```

- Gsynth: designed to handle block and staggered DID settings without treatment reversal, whereas other methods allow for treatment reversal under the assumption of limited carryover effects.
- Gsynth: particularly suited for cases where the number of treated units is small, including scenarios with only one treated unit. In contrast, other methods rely on large samples, particularly a large number of treated units, to obtain reliable standard errors and confidence intervals 
- Compared with IFEct (method = "ife"), Gsynth does not rely on pre-treatment data from the treated units to impute. This approach significantly speeds up computation and improves stability.
- Therefore, we recommend setting method = "gsynth" in fect for scenarios where the treatment does not reverse (or is coded accordingly) and the number of treated units is small .


## Augmented Synthetic Control (augsynth)

```{r}
asyn <- augsynth(Y ~ D | X1 + X2, id, time, simdata1)
print(asyn$nu)
asyn.summ <- summary(asyn)
plot(asyn.summ, levels = "Average")
```

## Trajectory Balancing

```{r}
out.tjbal <- tjbal(Y ~ D,
    data = simdata1,
    index = c("id", "time"), Y.match.npre = 0,
    demean = TRUE, vce = "boot", nsims = 200,
)
print(out.tjbal)
plot(out.tjbal)
```
