---
title: Quant II
subtitle: Lab 11
author: Yinxuan Wang
date: '2026-04-30'
execute:
  echo: true
  warning: false
  message: false
  cache: true
---

# Today's Agenda

- Multiple hypothesis testing 
- Missing data
- Differential measurement error
- Partial identification

## Multiple Testing Adjustments

- Test a family of $M$ hypotheses in one analysis
- **FWER** family-wise error rate
  - P(at least one false rejection) $< \alpha$
- **FDR** false discovery rate
  - E[false rejections/total rejections] $\leq q$
- FWER control is more stringent than FDR control

## FWER control methods recap

- **Bonferroni**: 
  - test each hypothesis at $\alpha/M$, or multiply each p-value by $M$
- **Holm**: 
  - Step-down Bonferroni
  - Controls FWER under arbitrary dependence and is at least as powerful as Bonferroni
- **Romano-Wolf**: 
  - Step-down resampling method
  - Uses the joint distribution of test statistics
  - Can be more powerful than Holm when outcomes are correlated.

## FDR control methods Recap

- **Benjamini-Hochberg**: 
  - controls FDR under independence or positive dependence
- **Benjamini-Yekutieli**: 
  - Controls FDR under arbitrary dependence 
  - but pays a large power cost

## A 10-outcome example

Simulate one experiment with ten outcomes

- `Y1` to `Y4` are true nulls
- `Y5` to `Y10` have nonzero treatment effects
- Some outcomes are correlated because they share latent factors
- Some effects are very small, so the adjustment methods can disagree

```{r}
library(tidyverse)
library(ggplot2)
library(fixest)
library(RColorBrewer)
library(kableExtra)
# devtools::install_github("s3alfisc/wildrwolf")
library(wildrwolf)

set.seed(1234)

simulate_10_outcomes <- function(N, seed = NULL) {
  if (!is.null(seed)) set.seed(seed)

  D <- rbinom(N, 1, 0.5)

  g <- rnorm(N)
  m <- rnorm(N)
  v <- rnorm(N)
  eps <- matrix(rnorm(N * 10), nrow = N)

  data.frame(
    id = 1:N,
    D = D,

    # Four true nulls
    Y1 = eps[, 1],
    Y2 = 0.60 * g + eps[, 2],
    Y3 = 0.60 * g + 0.60 * m + eps[, 3],
    Y4 = 0.60 * g + 0.60 * m + eps[, 4],

    # Six non-null outcomes
    Y5  = 0.65 * D + 0.60 * g + 0.60 * m + eps[, 5],
    Y6  = 0.55 * D + 0.60 * g + 0.60 * m + eps[, 6],
    Y7  = 0.45 * D + 0.55 * g + 0.55 * v + eps[, 7],
    Y8  = 0.35 * D + 0.55 * g + 0.55 * v + eps[, 8],
    Y9  = 0.30 * D + 0.40 * g + eps[, 9],
    Y10 = 0.25 * D + 0.40 * m + eps[, 10]
  )
}

dat <- simulate_10_outcomes(N=400, seed=1234)
outcomes <- paste0("Y", 1:10)
```

### Estimate the model using `fixest`

The `c(Y1, ..., Y10) ~ D` syntax estimates the same right-hand-side model for multiple outcomes.

The result is a `fixest_multi` object.
```{r}
models <- feols(
  c(Y1, Y2, Y3, Y4, Y5, Y6, Y7, Y8, Y9, Y10) ~ D,
  data = dat
)

models
```

```{r}
models_list <- as.list(models)
names(models_list) <- outcomes

extract_fixest_result <- function(model, outcome, term = "D") {
  ct <- coeftable(model)

  data.frame(
    outcome = outcome,
    estimate = ct[term, "Estimate"],
    se = ct[term, "Std. Error"],
    t_stat = ct[term, "t value"],
    p_raw = ct[term, "Pr(>|t|)"]
  )
}

res_unadjusted <- lapply(outcomes, function(y) {
  extract_fixest_result(models_list[[y]], outcome = y, term = "D")
}) %>%
  bind_rows() %>%
  mutate(
    truth = ifelse(outcome %in% paste0("Y", 1:4), "true null", "non-null")
  )

res_unadjusted %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  kable(caption = "Unadjusted treatment-effect estimates from ten fixest models")
```
### Apply p-value adjustments
```{r}
# wildrwolf
# Romano-Wolf adjusted p-values
# pass the list of hypotheses/models,
rw <- rwolf(
  models = models,
  param = "D",
  B = 999
)
rw

rw_df <- as.data.frame(rw)

# other adjustment methods
res_adjusted <- res_unadjusted %>%
  mutate(
    p_bonferroni = p.adjust(p_raw, method = "bonferroni"),
    p_holm = p.adjust(p_raw, method = "holm"),
    p_bh = p.adjust(p_raw, method = "BH"),
    p_by = p.adjust(p_raw, method = "BY"),
    p_romano_wolf = rw_df[["RW Pr(>|t|)"]]
  )

res_adjusted %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  kable(caption = "Raw and adjusted p-values")
```

### Power
```{r}
#| echo: false
#| results: hide
## Power tradeoffs across sample size
outcomes <- paste0("Y", 1:10)

outcome_formula <- as.formula(
  paste0("c(", paste(outcomes, collapse = ", "), ") ~ D")
)

fit_10_models <- function(dat) {
  feols(outcome_formula, data = dat)
}

adjust_pvalues <- function(models, B_rw = 999) {
  models_list <- as.list(models)
  names(models_list) <- outcomes

  raw_res <- lapply(outcomes, function(y) {
    ct <- coeftable(models_list[[y]])

    data.frame(
      outcome = y,
      truth = ifelse(y %in% paste0("Y", 1:4), "true null", "non-null"),
      estimate = ct["D", "Estimate"],
      se = ct["D", "Std. Error"],
      t_stat = ct["D", "t value"],
      p_raw = ct["D", "Pr(>|t|)"]
    )
  }) %>%
    bind_rows()
  
  capture.output({
    rw_df <- rwolf(
      models = models,
      param = "D",
      B = B_rw
    ) %>%
      as.data.frame()
  })

  rw_p_col <- grep("^RW", names(rw_df), value = TRUE)

  if (length(rw_p_col) != 1) {
    stop(
      "Could not uniquely identify the Romano-Wolf p-value column. Columns are: ",
      paste(names(rw_df), collapse = ", ")
    )
  }

  raw_res %>%
    mutate(
      p_bonferroni = p.adjust(p_raw, method = "bonferroni"),
      p_holm = p.adjust(p_raw, method = "holm"),
      p_bh = p.adjust(p_raw, method = "BH"),
      p_by = p.adjust(p_raw, method = "BY"),
      p_romano_wolf = rw_df[[rw_p_col]]
    )
}
summarize_rejections <- function(res, N, sim_id, alpha = 0.05) {
  res %>%
    select(
      outcome,
      truth,
      p_raw,
      p_bonferroni,
      p_holm,
      p_romano_wolf,
      p_bh,
      p_by
    ) %>%
    pivot_longer(
      cols = starts_with("p_"),
      names_to = "method",
      values_to = "p_value"
    ) %>%
    mutate(
      method = recode(
        method,
        p_raw = "Unadjusted",
        p_bonferroni = "Bonferroni",
        p_holm = "Holm",
        p_romano_wolf = "Romano-Wolf",
        p_bh = "BH/FDR",
        p_by = "BY/FDR"
      ),
      rejected = p_value <= alpha,
      is_null = truth == "true null",
      is_nonnull = truth == "non-null"
    ) %>%
    group_by(method) %>%
    summarize(
      N = N,
      sim_id = sim_id,
      any_false_rejection = any(rejected & is_null),
      false_discovery_count = sum(rejected & is_null),
      any_true_rejection = any(rejected & is_nonnull),
      true_rejection_count = sum(rejected & is_nonnull),
      nonnull_count = sum(is_nonnull),
      average_true_positive_rate = true_rejection_count / nonnull_count,
      .groups = "drop"
    )
}
run_one_power_sim <- function(N, sim_id, B_rw = 199, alpha = 0.05) {
  dat <- simulate_10_outcomes(
    N = N,
    seed = 100000 + 1000 * N + sim_id
  )

  models <- fit_10_models(dat)

  adjust_pvalues(models, B_rw = B_rw) %>%
    summarize_rejections(
      N = N,
      sim_id = sim_id,
      alpha = alpha
    )
}
```

```{r}
N_grid <- seq(100,500,100)

S <- 100
B_rw <- 199

power_sim <- lapply(N_grid, function(N_now) {
  lapply(seq_len(S), function(s) {
    run_one_power_sim(
      N = N_now,
      sim_id = s,
      B_rw = B_rw,
      alpha = 0.05
    )
  }) %>%
    bind_rows()
}) %>%
  bind_rows()
```

```{r}
#| echo: false
power_summary <- power_sim %>%
  group_by(N, method) %>%
  summarize(
    familywise_error_rate = mean(any_false_rejection),
    any_discovery_power = mean(any_true_rejection),
    average_power = mean(average_true_positive_rate),
    mean_false_discoveries = mean(false_discovery_count),
    mean_true_discoveries = mean(true_rejection_count),
    .groups = "drop"
  ) %>%
  mutate(
    method = factor(
      method,
      levels = c(
        "Unadjusted",
        "Bonferroni",
        "Holm",
        "Romano-Wolf",
        "BH/FDR",
        "BY/FDR"
      )
    )
  )
```

```{r}
#| echo: false
power_summary %>%
  select(N, method, any_discovery_power, average_power) %>%
  pivot_longer(
    cols = c(any_discovery_power, average_power),
    names_to = "power_type",
    values_to = "power"
  ) %>%
  mutate(
    power_type = recode(
      power_type,
      any_discovery_power = "At least one real effect rejected",
      average_power = "Share of real effects rejected"
    )
  ) %>%
  ggplot(aes(x = N, y = power, color = method, shape = method)) +
  geom_line() +
  geom_point(size = 2) +
  facet_wrap(~ power_type) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_color_brewer(palette = "Set1") +
  labs(
    x = "Sample size",
    shape = "Method",
    color = "Method"
  ) + theme_bw()
```



```{r}
#| echo: false
power_summary %>%
  ggplot(aes(x = N, y = familywise_error_rate, color = method, shape = method)) +
  geom_hline(yintercept = 0.05, linetype = "dashed") +
  geom_line() +
  geom_point(size = 2) +
  scale_color_brewer(palette = "Set1") +
  scale_y_continuous(limits = c(0, 0.3)) +
  labs(
    title = "False Positive Rate. Dashed line is 0.05.",
    x = "Sample size",
    color = "Method",
    shape = "Method"
  ) + theme_bw()
```


## Missing Data

- **MCAR**: in expectation the estimates are unbiased 
- **MAR**: can use covariates that determine missingness to impute the missing values of the outcome

### Missing outcome data in RCT

- Treatment is randomized
- Some people didn't respond
- Complete-case analysis estimates the effect among respondents 
  - not necessarily the full sample

### Complete-case estimates and Manski bounds

- Manski bounds use only bounded outcomes
- Here outcomes are in `[0, 1]`
- Missing treated outcomes are set to either 0 or 1
- Missing control outcomes are set to the opposite extreme

```{r}
#| echo: true
inv_logit <- function(x) 1 / (1 + exp(-x))
clip01 <- function(x) pmin(pmax(x, 0), 1)

simulate_missing_outcomes <- function(N = 5000, mechanism = "none", seed = 123) {
  set.seed(seed)

  X <- rnorm(N)
  D <- rbinom(N, 1, 0.5)

  Y0 <- clip01(0.35 + 0.18 * X + rnorm(N, 0, 0.1))
  Y1 <- clip01(Y0 + 0.12 + 0.05 * X + rnorm(N, 0, 0.06))
  Y <- ifelse(D == 1, Y1, Y0)

  p_R <- switch(
    mechanism,
    none = rep(1, N),
    mcar = rep(0.75, N),
    covariate = inv_logit(0.5 - 0.8 * abs(X)),
    outcome = inv_logit(0.5 - 3 * (Y - mean(Y))),
    treatment_outcome = inv_logit(0.5 + 3 * D - 3 * (Y - mean(Y))),
    stop("Unknown mechanism")
  )

  if (mechanism == "none") {
    R <- rep(1, N)
  } else {
    R <- rbinom(N, 1, p_R)
  }

  Y_obs <- Y
  Y_obs[R == 0] <- NA

  data.frame(
    id = 1:N,
    mechanism = mechanism,
    X = X,
    D = D,
    Y0 = Y0,
    Y1 = Y1,
    Y = Y,
    R = R,
    Y_obs = Y_obs
  )
}

estimate_missing_outcomes <- function(dat) {
  p1 <- mean(dat$R[dat$D == 1])
  p0 <- mean(dat$R[dat$D == 0])

  mu1 <- mean(dat$Y_obs[dat$D == 1], na.rm = TRUE)
  mu0 <- mean(dat$Y_obs[dat$D == 0], na.rm = TRUE)

  data.frame(
    mechanism = unique(dat$mechanism),
    true_ate = mean(dat$Y1 - dat$Y0),
    complete_case = mu1 - mu0,
    response_treated = p1,
    response_control = p0,
    manski_lower = mu1 * p1 - mu0 * p0 - (1 - p0),
    manski_upper = mu1 * p1 + (1 - p1) - mu0 * p0
  )
}

mechanisms <- c("none", "mcar", "covariate", "outcome", "treatment_outcome")

res_missing <- lapply(mechanisms, function(m) {
  simulate_missing_outcomes(mechanism = m, seed = 1234) %>%
    estimate_missing_outcomes()
}) %>%
  bind_rows() %>%
  mutate(
    bias = complete_case - true_ate
  )

res_missing %>%
  mutate(across(where(is.numeric), ~ round(.x, 3))) %>%
  kable(caption = "Missing outcomes - same experiment, different response mechanisms")
```

### Lee Bounds

- Provide bounds on the causal effect even with unbounded outcome
- Relies on monotonicity assumption similar to that used for LATE identification
- But in this case it serves to solve sample selection

- Treated respondent group contains
  - Always responders
  - Treatment-induced responders
- Control respondent group contains
  - Always responders
- Bound for the ATE for the always responder sub-population

```{r}
# devtools::install_github("vsemenova/leebounds") 
# try devtools::install_github("haluong89-bcn/leebounds") if the vsemenova/leebounds doesn't work
library(leebounds)

dat_lee <- simulate_missing_outcomes(
  N = 5000,
  mechanism = "treatment_outcome",
  seed = 1234
)

lee_input <- dat_lee %>%
  transmute(
    treat = D,
    outcome = Y_obs,
    selection = R
  )

lee_raw <- leebounds(lee_input)

do.call("rbind", lee_raw)
```

### Distribution of complete-case estimates with different missing mechanism

```{r}
#| echo: false
mechanisms <- c("mcar", "covariate", "outcome", "treatment_outcome")
one_missing_sim <- function(mechanism, sim_id, N) {
  simulate_missing_outcomes(
    N = N,
    mechanism = mechanism,
    seed = 1000 + sim_id
  ) %>%
    estimate_missing_outcomes() %>%
    mutate(
      sim_id = sim_id,
      bias = complete_case - true_ate
    )
}
missing_mc <- lapply(mechanisms, function(m) {
  lapply(1:500, function(s) {
    one_missing_sim(mechanism = m, sim_id = s, N = 500)
  }) %>%
    bind_rows()
}) %>%
  bind_rows()
missing_mc %>%
  mutate(
    mechanism = factor(
      mechanism,
      levels = mechanisms,
      labels = c(
        "MCAR",
        "Covariate",
        "Outcome",
        "Treatment_Outcome"
      )
    )
  ) %>%
  ggplot(aes(x = complete_case)) +
  geom_density() +
  geom_vline(
    data = missing_mc %>%
      group_by(mechanism) %>%
      summarize(mean_true_ate = mean(true_ate), .groups = "drop") %>%
      mutate(
        mechanism = factor(
          mechanism,
          levels = mechanisms,
          labels = c(
            "MCAR",
            "Covariate",
            "Outcome",
            "Treatment_Outcome"
          )
        )
      ),
    aes(xintercept = mean_true_ate),
    linetype = "dashed"
  ) +
  facet_wrap(~ mechanism, ncol = 2) +
  labs(
    title = "Distribution of complete-case estimates. Dashed line is the average true ATE.") +
  theme_bw()
```


### Missing data in confounders in observational studies
```{r}
set.seed(1234)

sim_once <- function(n = 3000, tau = 1, missing = c("MAR", "MNAR")) {
  missing <- match.arg(missing)
  
  # two pre-treatment confounders
  X1 <- rnorm(n)
  X2 <- 0.6 * X1 + sqrt(1 - 0.6^2) * rnorm(n)
  
  p_D <- plogis(-0.2 + X1 + X2)
  D <- rbinom(n, 1, p_D)
  
  Y <- tau * D + 1 * X1 + 1.5 * X2 + rnorm(n)
  
  # Missingness in X2
  # MAR: missingness depends only on observed X1.
  # MNAR: missingness depends on X2 itself.
  if (missing == "MAR") {
    p_obs <- plogis(1 - 2 * X1)
  } else {
    p_obs <- plogis(1 - 2 * X2)
  }
  
  # R = 1 means X2 observed
  R <- rbinom(n, 1, p_obs)
  X2_obs <- ifelse(R == 1, X2, NA)
  
  dat <- data.frame(Y, D, X1, X2, X2_obs, R)
  
  # Estimators
  # Oracle: what we would estimate if X2 were fully observed.
  oracle <- coef(lm(Y ~ D + X1 + X2, data = dat))["D"]
  
  # Naive: omit the missing confounder entirely.
  omit_x2 <- coef(lm(Y ~ D + X1, data = dat))["D"]
  
  # Complete case: use only units where X2 is observed.
  complete_case <- coef(
    lm(Y ~ D + X1 + X2_obs, data = dat, subset = R == 1)
  )["D"]
  
  # Mean imputation 
  dat$X2_mean_imp <- ifelse(is.na(X2_obs), mean(X2_obs, na.rm = TRUE), X2_obs)
  mean_imp <- coef(
    lm(Y ~ D + X1 + X2_mean_imp, data = dat)
  )["D"]
  
  # Missingness-indicator method with Lin (2013) fully interacted
  dat$X1_c <- dat$X1 - mean(dat$X1)
  dat$X2_c <- dat$X2 - mean(dat$X2)
  dat$M2 <- as.integer(is.na(dat$X2_obs))
  dat$X2_c[dat$M2 == 1] <- 0
  mim_lin <- coef(
    lm(Y ~ D * (X1_c + X2_c + M2), data = dat)
  )["D"]
  
  data.frame(
    missing = missing,
    estimator = c(
      "Oracle, X2 fully observed",
      "Omit X2",
      "Complete case",
      "Mean imputation",
      "Impute 0 + missingness indicator"
    ),
    estimate = c(
      oracle,
      omit_x2,
      complete_case,
      mean_imp,
      mim_lin
    ),
    true_effect = tau
  )
}

# One draw
rbind(
  sim_once(missing = "MAR"),
  sim_once(missing = "MNAR")
)
```

```{r}
run_mc <- function(B = 200, missing = c("MAR", "MNAR")) {
  missing <- match.arg(missing)
  
  out <- do.call(
    rbind,
    lapply(seq_len(B), function(b) {
      res <- sim_once(missing = missing)
      res$sim <- b
      res
    })
  )
  
  aggregate(
    estimate ~ missing + estimator,
    data = out,
    FUN = function(x) c(mean = mean(x), sd = sd(x))
  )
}

mc_mar <- run_mc(B = 200, missing = "MAR")
mc_mnar <- run_mc(B = 200, missing = "MNAR")

mc <- rbind(mc_mar, mc_mnar)
mc <- do.call(data.frame, mc)
names(mc) <- c("missing", "estimator", "mean_estimate", "sd_estimate")
mc$bias <- mc$mean_estimate - 1

mc
```

```{r}
ggplot(mc, aes(x = estimator, y = mean_estimate)) +
  geom_hline(yintercept = 1, linetype = "dashed") +
  geom_point(size = 2) +
  geom_errorbar(
    aes(
      ymin = mean_estimate - 1.96 * sd_estimate,
      ymax = mean_estimate + 1.96 * sd_estimate
    ),
    width = 0.15
  ) +
  facet_wrap(~ missing) +
  coord_flip() +
  labs(
    x = NULL,
    subtitle = "Dashed line is the true treatment effect"
  ) +
  theme_bw()
```

####  Some recent/current work on missing data 
<https://naijialiu.github.io>

## Differential Measurment Errors

<https://imai.fas.harvard.edu/research/files/merror.pdf>

```{r}
set.seed(1234)

sim_merror_once <- function(
  n = 5000,
  tau = 0.5,
  gamma = 0
) {
  
  Z_star <- rbinom(n, 1, 0.5)
  
  Y <- tau * Z_star + rnorm(n)
  
  # gamma = 0:
  #   reporting depends only on true treatment.
  #   This is nondifferential measurement error.
  # gamma > 0: Differential measurement error.
  #   high-Y units are more likely to be reported as treated.
  #   low-Y units are more likely to be reported as control.
  p_report_treated <- plogis(-2 + 4 * Z_star + gamma * Y)
  Z_obs <- rbinom(n, 1, p_report_treated)
  
  oracle <- coef(lm(Y ~ Z_star))["Z_star"]
  naive  <- coef(lm(Y ~ Z_obs))["Z_obs"]
  
  data.frame(
    gamma = gamma,
    oracle = oracle,
    naive = naive,
    true_effect = tau,
    bias_naive = naive - tau)
}

rbind(
  sim_merror_once(gamma = 0),
  sim_merror_once(gamma = 0.5),
  sim_merror_once(gamma = 1.0),
  sim_merror_once(gamma = 1.5),
  sim_merror_once(gamma = 2.0)
)
```

```{r}
sim_merror_once <- function(n = 500, tau = 0.5, gamma = 0) {
  Z_star <- rbinom(n, 1, 0.5)
  Y <- tau * Z_star + rnorm(n)
  
  p_report_treated <- plogis(-2 + 4 * Z_star + gamma * Y)
  Z_obs <- rbinom(n, 1, p_report_treated)
  
  data.frame(
    oracle = coef(lm(Y ~ Z_star))["Z_star"],
    naive  = coef(lm(Y ~ Z_obs))["Z_obs"]
  )
}

mc_merror <- expand.grid(
  gamma = seq(0, 2, by = 0.25),
  sim = 1:200
) %>%
  pmap_dfr(function(gamma, sim) {
    sim_merror_once(gamma = gamma) %>%
      mutate(gamma = gamma, sim = sim)
  }) %>%
  pivot_longer(
    cols = c(oracle, naive),
    names_to = "estimator",
    values_to = "estimate"
  ) %>%
  group_by(gamma, estimator) %>%
  summarise(
    mean = mean(estimate),
    se = sd(estimate) / sqrt(n()),
    .groups = "drop"
  )

ggplot(mc_merror, aes(x = gamma, y = mean, linetype = estimator)) +
  geom_hline(yintercept = 0.5, linetype = "dashed") +
  geom_ribbon(
    aes(ymin = mean - 1.96 * se, ymax = mean + 1.96 * se, group = estimator),
    alpha = 0.15,
    linetype = 0
  ) +
  geom_line() +
  geom_point() +
  labs(
    x = "Strength of outcome-dependent reporting error",
    y = "Estimated treatment effect",
    subtitle = "Dashed line is the true effect; bands are 95% Monte Carlo SE intervals",
    linetype = NULL
  ) +
  theme_bw()
```

## Some recent/current work Partial Identification

[Autobounds](https://jmummolo.scholar.princeton.edu/sites/g/files/toruqf3341/files/documents/autobounds_general_draft_0.pdf)

The computational idea behind AutoBounds:

1. Define the causal estimand.
2. Specify the latent causal/data-generating process.
3. Specify which variables or margins are observed in the discrete data.
4. Translate probability axioms, observed-data restrictions, and assumptions into equality or inequality constraints.
5. Search over all latent distributions or DGPs consistent with those constraints.
6. Minimize and maximize the estimand over that feasible set.

Minimum $=$ maximum, the estimand is point identified.

Minimum $\neq$ maximum, sharp bounds.

### selection into administrative data example

Racial bias in police use of force using police records

- Treatment $D_i$: civilian race
- Mediator $M_i$: whether the police stop the civilian
- Outcome $Y_i$: force used (zero if no stop)
- Unobserved $U_i$: e.g., officer suspicion / sense of threat

[Autobounds demo by Cooper Duarte and Knox 2023](https://github.com/duarteguilherme/autobounds/blob/main/autobounds_demo/apsa2023/klm_demo.ipynb)
