---
title: Quant II
subtitle: Lab 9
author: Yinxuan Wang
date: '2026-04-16'
execute:
  echo: true
  warning: false
  message: false
  cache: true
---



## Today's Agenda
- Experiment
- DeclareDesign 
  - adapted from <https://declaredesign.org>
- Covariate adjustment

## Purpose of `DeclareDesign`
Simulation based toolkit

- make assumptions explicit,
- define the estimand clearly,
- declare how data are generated,
- declare how the answer will be computed,
- diagnose whether the design performs well,
- redesign when it does not.

## `DecalreDesign` Workflow
1. start with the research question,
2. write down the causal model,
3. declare the estimand,
4. declare assignment and measurement,
5. declare the estimator,
6. diagnose bias, RMSE, coverage, and power,
7. redesign over the choices we actually control.

## Setup

```{r}
pacman::p_load(DeclareDesign, tidyverse, future, future.apply)
library(fabricatr)
library(randomizr)
library(estimatr)

set.seed(12345)
theme_set(theme_minimal(base_size = 14))
```

# Baseline 

```{r}
N <- 300
tau <- 0.20
x_strength <- 1.00

base_dd <-
  declare_model(
    N = N,
    U = rnorm(N),
    potential_outcomes(Y ~ tau * Z + U)
  ) +
  declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  declare_assignment(Z = complete_ra(N = N, prob = 0.5)) +
  declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  declare_estimator(Y ~ Z, inquiry = "ATE", label = "DIM")

base_dd
```


# Diagnose the baseline design

```{r}
diagnosis_base_dd <- diagnose_design(base_dd)
```
```{r}
head(diagnosis_base_dd$simulations_df)
hist(diagnosis_base_dd$simulations_df$estimate)
hist(diagnosis_base_dd$simulations_df$p.value)
diagnosis_base_dd$diagnosands_df
```

<https://book.declaredesign.org/declaration-diagnosis-redesign/diagnosing-designs.html#sec-ch10s2>

## `redesign`: Evaluating Multiple Designs

### `redesign` over sample size

```{r}
diagnosis_over_n <-
  base_dd %>%
  redesign(N = seq(100, 1000, by = 100)) %>%
  diagnose_designs()
```
```{r}
diagnosis_over_n$diagnosands_df %>%
  select(N, estimator, power, sd_estimate, rmse)
```


```{r}
diagnosis_over_n$diagnosands_df %>% 
  ggplot(aes(x = N, y = mean_estimate, ymin = mean_estimate - 1.96 * sd_estimate, ymax = mean_estimate + 1.96 * sd_estimate)) +
  geom_point(size = 4) +
  geom_errorbar(width = 0, linewidth = 1) + 
  labs(
    title = "95% CI",
    x = "Sample size",
    y = "Estimate") + 
  theme_bw()

diagnosis_over_n$diagnosands_df %>%
  ggplot(aes(x = N, y = power, color = estimator)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  labs(
    title = "Power by sample size",
    x = "Sample size",
    y = "Power") + 
  theme_bw() +
  geom_hline(yintercept = 0.8, linetype = "dashed") +
  annotate("text", x = Inf, y = 0.8, label = "Power = 0.8", hjust = 1.1, vjust = 1.5)
```

### `redesign` the effect size

```{r}
diagnosis_over_tau <- 
  base_dd %>%
    redesign(
        N = seq(100, 1000, 100),
        tau = c(0.1, 0.2, 0.3)
    ) %>%
    diagnose_designs()
```
```{r}
diagnosis_over_tau$diagnosands_df %>% 
  ggplot(aes(x = N, y = power, color = as.factor(tau))) +
  geom_line(linewidth = 2) +
  labs(color = "tau",
       title = "Power by effect size") + 
  theme_bw() +
  geom_hline(yintercept = 0.8, linetype = "dashed") +
  annotate("text", x = Inf, y = 0.8, label = "Power = 0.8", hjust = 1.1, vjust = -0.5)
```

## Covariate Adjustment

1. Difference in means
2. OLS with covariate adjustment
3. Lin adjustment using `lm_lin()`

```{r}
adj_dd <-
  declare_model(
    N = N,
    X = rnorm(N),
    U = rnorm(N),
    potential_outcomes(Y ~ tau * Z + x_strength * X + U)
  ) +
  declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  declare_assignment(Z = complete_ra(N = N, prob = 0.5)) +
  declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  declare_estimator(Y ~ Z, inquiry = "ATE", label = "DIM") +
  declare_estimator(Y ~ Z + X, .method = lm_robust, inquiry = "ATE", label = "OLS adjusted") +
  declare_estimator(Y ~ Z, covariates = ~ X, .method = lm_lin, inquiry = "ATE", label = "Lin")
```

### `redesign` over covariate strength

```{r}
diagnosis_adj <-
  adj_dd %>%
  redesign(
    N = seq(100, 1000, by = 100),
    x_strength = c(0, 0.5, 1.0, 1.5)
  ) %>%
  diagnose_designs()
```
```{r}
diagnosis_adj$diagnosands_df %>%
  ggplot(aes(x = N, y = sd_estimate, color = estimator)) +
  geom_line(linewidth = 1.1) +
  facet_wrap(~ x_strength, labeller = label_both) +
  labs(
    title = "Precision gains from covariate adjustment",
    x = "Sample size",
    y = "SD of estimate"
  ) + theme_bw()
```
```{r}
diagnosis_adj$diagnosands_df %>%
  ggplot(aes(x = N, y = power, color = estimator)) +
  geom_line(linewidth = 1.1) +
  facet_wrap(~ x_strength, labeller = label_both) +
  labs(
    x = "Sample size",
    y = "Power"
  ) + 
  theme_bw() +
  geom_hline(yintercept = 0.8, linetype = "dashed") +
  annotate("text", x = Inf, y = 0.8, label = "Power = 0.8", hjust = 1.1, vjust = -0.5)
```


### When OLS is less efficient than Lin

- Covariate slopes differ between treated and control potential outcomes, 
- and assignment is imbalanced

```{r}
N <- 400
tau <- 0.2
treatment_share <- 0.5
control_slope <- 0
treatment_slope <- 1

heterogeneous_slope_dd <-
  declare_model(
    N = N,
    X = runif(N, min = -1, max = 1),
    U = rnorm(N, sd = 0.25),
    Y_Z_0 = control_slope * X + U,
    Y_Z_1 = tau + treatment_slope * X + U
  ) +
  declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  declare_assignment(Z = complete_ra(N = N, prob = treatment_share)) +
  declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  declare_estimator(Y ~ Z, inquiry = "ATE", label = "DIM") +
  declare_estimator(Y ~ Z + X, .method = lm_robust, inquiry = "ATE", label = "OLS adjusted") +
  declare_estimator(Y ~ Z, covariates = ~ X, .method = lm_lin, inquiry = "ATE", label = "Lin")
```

```{r}
heterogeneous_slope_diagnosis <-
  heterogeneous_slope_dd %>%
  redesign(
    treatment_share = c(0.20, 0.35, 0.50),
    control_slope = c(-1, 0, 1)
  ) %>%
  diagnose_designs()
```

```{r}
heterogeneous_slope_diagnosis$diagnosands_df |>
  ggplot(aes(x = treatment_share, y = sd_estimate, color = estimator)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  facet_wrap(~ control_slope, labeller = label_both) +
  labs(
    title = "Estimator precision under slope heterogeneity and imbalance",
    x = "Treatment share",
    y = "SD of estimate") +
  theme_bw()
```

```{r}
heterogeneous_slope_diagnosis$diagnosands_df |>
  ggplot(aes(x = treatment_share, y = power, color = estimator)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  facet_wrap(~ control_slope, labeller = label_both) +
  labs(
    x = "Treatment share",
    y = "Power") +
  theme_bw() +
  geom_hline(yintercept = 0.8, linetype = "dashed") +
  annotate("text", x = Inf, y = 0.8, label = "Power = 0.8", hjust = 1.1, vjust = 1.5)
```


## Block Randomization
```{r}
block_N <- 200
tau <- 0.2
coef <- 2

block_dd <-
  declare_model(
    N = block_N,
    gender = sample(c(0, 1), size = N, replace = TRUE),
    U = rnorm(N, sd = 1),
    potential_outcomes(Y ~ tau * Z + coef * gender + U)
  ) +
  declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  declare_assignment(
    Z = block_ra(blocks = gender)
  ) +
  declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  declare_estimator(Y ~ Z, inquiry = "ATE", label = "DIM") +
  declare_estimator(
    Y ~ Z,
    covariates = ~ gender,
    .method = lm_lin,
    inquiry = "ATE",
    label = "Lin"
  )
```
```{r}
simple_dd <-
  declare_model(
    N = block_N,
    gender = sample(c(0, 1), size = N, replace = TRUE),
    U = rnorm(N, sd = 1),
    potential_outcomes(Y ~ tau * Z + coef * gender + U)
  ) +
  declare_inquiry(ATE = mean(Y_Z_1 - Y_Z_0)) +
  declare_assignment(
    Z = complete_ra(N = N, prob = 0.5)
  ) +
  declare_measurement(Y = reveal_outcomes(Y ~ Z)) +
  declare_estimator(Y ~ Z, inquiry = "ATE", label = "DIM") +
  declare_estimator(
    Y ~ Z,
    covariates = ~ gender,
    .method = lm_lin,
    inquiry = "ATE",
    label = "Lin"
  )
```

```{r}
diagnosis_block <-
  block_dd %>%
  redesign(
    block_N = seq(40, 200, 20),
    coef = c(0.5, 1, 2, 4)
  ) %>%
  diagnose_designs()
```

```{r}
diagnosis_simple <-
  simple_dd %>%
  redesign(
    block_N = seq(40, 200, 20),
    coef = c(0.5, 1, 2, 4)
  ) %>%
  diagnose_designs()
```

```{r}
bind_rows(
  diagnosis_block$diagnosands_df %>% mutate(design = "Blocked assignment"),
  diagnosis_simple$diagnosands_df %>% mutate(design = "Complete randomization")
) %>%
  ggplot(aes(x = block_N, y = sd_estimate, color = design)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ estimator + coef, ncol = 4) +
  labs(
    x = "Sample size",
    y = "SD of estimate",
    color = "Design"
  ) +
  theme_bw()
```

```{r}
bind_rows(
  diagnosis_block$diagnosands_df %>% mutate(design = "Blocked assignment"),
  diagnosis_simple$diagnosands_df %>% mutate(design = "Complete randomization")
) %>%
  ggplot(aes(x = block_N, y = power, color = design)) +
  geom_line(linewidth = 1) +
  facet_wrap(~ estimator + coef, ncol = 4) +
  labs(
    x = "Sample size",
    y = "Power",
    color = "Design"
  ) +
  theme_bw() +
  geom_hline(yintercept = 0.8, linetype = "dashed") +
  annotate("text", x = Inf, y = 0.8, label = "Power = 0.8", hjust = 1.1, vjust = 1.5)
```


## Factorial Design 
```{r}
N <- 800
tau_A <- 0.20
tau_B <- 0.20
tau_AB <- 0.20

factorial_dd <-
  declare_model(
    N = N,
    X = rnorm(N),
    U = rnorm(N),
    Y_A_0_B_0 = 0 + U,
    Y_A_1_B_0 = tau_A + U,
    Y_A_0_B_1 = tau_B + U,
    Y_A_1_B_1 = tau_A + tau_B + tau_AB + U
  ) +
  declare_inquiry(
    main_A = mean((Y_A_1_B_0 + Y_A_1_B_1)/2 - (Y_A_0_B_0 + Y_A_0_B_1)/2),
    main_B = mean((Y_A_0_B_1 + Y_A_1_B_1)/2 - (Y_A_0_B_0 + Y_A_1_B_0)/2),
    interaction = mean((Y_A_1_B_1 - Y_A_0_B_1) - (Y_A_1_B_0 - Y_A_0_B_0))
  ) +
  declare_assignment(
    A = complete_ra(N = N, prob = 0.5),
    B = complete_ra(N = N, prob = 0.5)
  ) +
  declare_measurement(
    Y = reveal_outcomes(Y ~ A + B)
  ) +
  declare_estimator(
    Y ~ A + B + A:B,
    model = lm_robust,
    inquiry = c("main_A", "main_B", "interaction"),
    term = c("A", "B", "A:B"),
    label = "OLS factorial"
  )
```

```{r}
factorial_diag <-
  factorial_dd %>%
  redesign(N = seq(500, 7000, by = 500)) %>%
  diagnose_designs()
```

```{r}
factorial_diag$diagnosands_df %>%
  ggplot(aes(x = N, y = power, color = term)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2) +
  labs(
    title = "Power in a 2x2 factorial design",
    x = "Sample size",
    y = "Power",
    color = "Estimand"
  ) +
  theme_bw() +
  geom_hline(yintercept = 0.8, linetype = "dashed") +
  annotate("text", x = Inf, y = 0.8, label = "Power = 0.8", hjust = 1.1, vjust = -0.5)
```

## DeclareDesign in Observatinal Settings Example
```{r}
N <- 500
tau <- 0.5
selection_strength <- 1.2
confounding_strength <- 1.0

obs_dd <-
  declare_model(
    N = N,
    X = rnorm(N),
    U = rnorm(N),
    D = rbinom(N, 1, plogis(-0.3 + selection_strength * X)),
    Y_D_0 = confounding_strength * X + U,
    Y_D_1 = Y_D_0 + tau
  ) +
  declare_inquiry(ATE = mean(Y_D_1 - Y_D_0)) +
  declare_measurement(Y = reveal_outcomes(Y ~ D)) +
  declare_estimator(Y ~ D,
                    .method = lm_robust,
                    inquiry = "ATE",
                    label = "Raw OLS") +
  declare_estimator(Y ~ D + X,
                    .method = lm_robust,
                    inquiry = "ATE",
                    label = "Adjusted OLS")
```

```{r}
obs_diag <- diagnose_design(obs_dd)
obs_diag$diagnosands_df
```

```{r}
obs_diag_grid <-
  obs_dd %>%
  redesign(
    selection_strength = c(0.4, 0.8, 1.2, 1.6),
    confounding_strength = c(0.5, 1.0, 1.5)
  ) %>%
  diagnose_designs()
```

```{r}
obs_diag_grid$diagnosands_df %>%
  ggplot(aes(x = selection_strength, y = bias, color = estimator)) +
  geom_line(linewidth = 1.1) +
  facet_wrap(~ confounding_strength, labeller = label_both) +
  labs(
    title = "Bias in an observational design",
    x = "How strongly X predicts treatment",
    y = "Bias"
  ) +
  theme_bw()
```
