---
title: Quant II
subtitle: Lab 3
author: Yinxuan Wang
date: '2026-02-19'
format:
  beamer:
    theme: CambridgeUS
    colortheme: dolphin
    fonttheme: structurebold
    incremental: false
---

# Homeworks

- Homework 1 grades before next recitation

Notes:

- Email to **both** Cyrus \& me
- Send **one PDF**, including math, code, and output 
  - Use `.qmd` or `.rmd` for code+output
  - Math part can be scanned and attached (or from tablet)\newline LaTeX recommended
- Deadlines are before the lecture


# Today's Agenda

- Covariate Adjustment in Experimental Data
- Multiple Regression
- Omitted Variable Bias
- Matching

#  Covariates for efficiency in randomized experiments
- Lin (2013), fully interacted regression
$$Y_i = \alpha + \rho D_i + (X_i - \bar{X})'\beta_0 + D_i(X_i-\bar{X})'\beta_1 + \epsilon_i$$
- Difference in residuzlied means: $\hat{\rho} = \bar{\tilde{Y_1}} - \bar{\tilde{Y_0}}$

- Efficiency gain: $\text{Var}(\tilde{Y_i}(d)) \approx \sigma^2_{Y(d)}(1-R^2_d)$

:::{.incremental}
- Intuition: reduce the unexplained variation of the outcome, reduce the standard error on the causal estimate
- In small samples instability may arise from issues like covariate imbalance across treatment arms, high leverage, low degree of freedoms
    - Need enough observations **per treatment arm**
    - Prefer a small, pre-specified covariate set strongly predictive of outcome
:::

# Implementation with `estimatr::lm_lin`
```{r}
#| echo: true
library(fabricatr); library(randomizr)

set.seed(123)

dat <- fabricate(N=1000, x=rnorm(N, mean=2.3),
                 x2=rpois(N, lambda = 2),
                 x3=runif(N),
                 y0=rnorm(N)+x,
                 y1=rnorm(N)+x+0.7)
dat$z <- complete_ra(N=nrow(dat))
dat$y <- ifelse(dat$z==1, dat$y1, dat$y0)

# lm_lin
out1 <- estimatr::lm_lin(y~z, covariates=~x, dat)
```

# Implementation, alternatives
```{r}
#| echo: true

# lm_robust
dat$x_c <- dat$x-mean(dat$x)
out2 <- estimatr::lm_robust(y~z*x_c, dat)

# manually
fit1 <- lm(y ~ x_c, dat[dat$z==1,])
fit0 <- lm(y ~ x_c, dat[dat$z==0,])
rho <- coef(fit1)["(Intercept)"] - coef(fit0)["(Intercept)"]

# compare all
c(coef(out1)["z"], coef(out2)["z"], rho)

```

# Strong Ignorability Assumptions
Given $(Y_i, X_i)$, the following allow $\rho$ to be identified:

$$D_i \perp (Y_{1i}, Y_{0i}) \quad \text{and} \quad
0 < \Pr[D_i = 1] < 1
\quad \text{for all } i \in U$$

. . . 

or

$$D_i \perp (Y_{1i}, Y_{0i}) \mid X_i
\quad \text{and} \quad
0 < \Pr[D_i = 1 \mid X_i = x] < 1
\quad \text{for all } x \in \mathcal{X}.$$

- Informally: ``selection on observables''
- Considered strong in most contexts

# Identification under CIA
Guarantees that covariate stratum-specific ATEs are identified in the data:
$$
\begin{aligned}
\tau(x) &= E[Y_{1i} - Y_{0i}|X_i=x] \\
& = E[Y_{1i}|D_i=1, X_i=x] - E[Y_{0i}|D_i=0, X_i=x]
\end{aligned}
$$

. . . 

And then $\color{red}\tau_{PATE}$ is identified by averaging over the covariates distribution (cf. Imbens and Wooldridge 2009, p.26-27)
$$
{\color{red} \tau_{PATE}} = E[\tau(x)] = \int_X\tau(x){\color{red}dF(x)}
$$

. . . 

Same holds for $\color{blue}\tau_{PATT}$, changing the distribution over which to average: 
$$
{\color{blue} \tau_{PATT}} = E[\tau(x)|D_i=1] = \int_{X|D=1}\tau(x){\color{blue}dF(x|D_i=1)}
$$

# Multiple regression as an identification strategy
> - Multiple regression is an obvious tool to estimate causal effects under CIA 
> - Estimates linear approximations of the conditional means of the potential outcomes

# Frisch-Waugh-Lovell theorem, a refresher
- Linear model with $k$ covariates. In matrix form: $y = X'\beta + \varepsilon$
- FWL gives a formula for the OLS estimate of the $k^{th}$ coefficient.

$$
\hat{\beta}_k = (X'_{k}M_{[X_{-k}]}X_{k})^{-1}X'_{k}M_{[X_{-k}]}y
$$

\pause Equivalent to:

:::{.incremental}
- Regress the $X_k$ on $X_{-k}$ and take the residuals
- Regress the $y$ on $X_{-k}$ and take the residuals
- Regress the residuals of $y$ on the residuals for $X_k$
    - Note: that to get $\hat{\beta}_k$ it is enough to regress the non-residualized $y$ on residualized $X_k$ (why?) 
    - but the SE won't be right
- We are often interested in just one regressor (e.g. treatment indicator)
:::


# FWL in practice
\small
```{r}
#| echo: true

library(tidyverse)
data("mtcars")

# multivariate regression
fit <- lm(mpg ~ cyl + drat + wt, mtcars)

# FWL
resy <- lm(mpg ~ drat + wt, mtcars) %>% residuals()
resx <- lm(cyl ~ drat + wt, mtcars) %>% residuals()
fit2 <- lm(resy ~ resx)

# compare results
out <- c(coefficients(fit)["cyl"], coefficients(fit2)["resx"])
names(out) <- c("Multivariate", "Univariate Residualized")
out
```

# Omitted Variable Bias, Intuition
Suppose the true outcome model has treatment $D_i$ and covariates $X_i$:
$$Y_i = \alpha + \rho D_i + X_i'\gamma + \nu_i$$ \pause
Decompose both variables into "explained by X" + "residual" pieces:
$$D_i = \widehat{E}[D_i\mid X_i] + \tilde D_i,\quad
Y_i = \widehat{E}[Y_i\mid X_i] + \tilde Y_i.$$ \pause
Because residuals are orthogonal to the regressors, cross-covariances are zero
$$\text{Cov}(Y_i,D_i)
= \text{Cov}(\widehat{E}[Y_i\mid X_i],\widehat{E}[D_i\mid X_i])\;+\;\text{Cov}(\tilde Y_i,\tilde D_i)$$ \pause

- FWL/controlling for $X$ removes the first component by construction
- Omitting $X$ leaves it in

# Omitted Variable Bias
The coefficient on $D_i$ is $\frac{\text{Cov}(Y_i,D_i)}{\text{Var}(D_i)}$

Omitting $X$,
$$\text{Cov}(Y_i, D_i) = \underbrace{\text{Cov}(\rho D_i + X_i'\gamma + v, D_i)}_{\text{Total Variation}}$$
Then, 
\begin{align*}
  \hat{\rho}_{short} & = \rho + \gamma' \cdot \frac{\text{Cov}(X_i, D_i)}{\text{Var}(D_i)} \\
  &= \rho + \gamma'\delta
\end{align*}

OVB: ``Confounder Impact $\times$ Imbalance'' (Cinelli \& Hazlett 2020)

# Sensitivity analysis, i
>- CIA is untestable
>- How results could be affected by hypothetical departures from it
>- Cinelli and Hazlett (2020), implemented through `sensemakr`
>- Package website: \url{https://carloscinelli.com/sensemakr/}

# Sensitivity analysis, ii
> - A full model should include covariates $X$ and $Z$, but instead
> - Estimate $Y = \rho_{partial} D + X'\beta + \varepsilon$
> - We know that $\hat{\rho}_{partial} = \rho_{true} + {{\gamma}{\delta}}$, where ${\delta}=\frac{Cov(\tilde{D}, \tilde{Z})}{V(\tilde{D})}$
> - We don't know $\gamma$ or $\delta$
> - Hypothesize different values for them and see if $\rho_{true}$ would cross 0.
> - e.g., "If $Z$ is as strong as $X$, does my effect disappear?"

# Sensitivity analysis example, i

**Attitudes in Darfur (Hazlett, 2019)**

- 2003-2004 government violence against civilians 
- Outcome ($Y$): attitudes toward peace
- Treatment ($D$): exposure to violence

```{r}
#| message: false
#| echo: true

library(sensemakr); library(fixest)
data("darfur")
darfur.model <- feols(
    peacefactor ~ directlyharmed + female +
        age + farmer_dar + herder_dar + pastvoted + 
        hhsize_darfur | village,
    data = darfur
)
```

# Sensitivity analysis example, ii
\small
```{r}
#| echo: true
etable(darfur.model)
```
# Sensitivity analysis example, iii
```{r}
#| echo: true
sensitivity <- sensemakr(darfur.model, 
                         treatment = "directlyharmed", 
                         benchmark_covariates = "female", 
                         kd = 1:4)

# summary(sensitivity)
```

> - `sensemakr` package lets us conduct sensitivity analysis 
      **relative to a covariate of choice**
> - `kd` and `ky` arguments: a hypothetical confounder **1-4x** as strong as `female`

# Sensitivity analysis example, iv
\tiny
```{r}
#| echo: true
#| fig-width: 4
#| fig-height: 3

plot(sensitivity)
```

# Other issues with multiple regression
- So far, we assumed $\rho_i = \rho \, \forall i$ and linearity
- Both can go wrong!
\pause

**When there are effect heterogeneity**

- Multiple regression estimates a conditional variance-weighted average 
- Weights are proportional to $Var(D_i|X_i)$
- Not necessarily $=$ the population causal effect.
\pause

**When linearity fails**

- If we omit interactions, higher order terms
- Specification error

Double machine learning can help (maybe next week?)
- but not transparent 

# Matching
General idea: For each treated unit, pick untreated unit(s) with similar covariate values.
\pause

- Non-parametric
- Transparent
- Estimation technique
- Doesn’t solve issues with bad/good controls, selection on unobservables, etc.

\pause
Then,

- How many units? 
- What counts as similar?

# Weaker identifying assumptions for ATT

**Conditional Mean Independence** (CMI)

$$E[Y_{0i}|D_i=1, X_i] = E[Y_{0i}|D_i=0, X_i], \quad P(D_i=1|X_i)<1$$

- After conditioning on $X$, treated and control units have the same expected outcome under control.
- Treated units have comparable controls

# Exact Matching

Choose matches that have exactly the same value of $X_i$

- For each treated unit, find control units with same values of $X_i$
- Compute the difference in means within these strata and 
- For ATT, compute a weighted average using the distribution of $X$ of the treated group
\pause

- Problems
  - Only works with discrete covariates
  - Overlap problems: no control unit available for some levels of $X_i$

# Exact matching, illustration

```{r}
#| echo: false
#| fig-width: 8
#| fig-height: 4
pacman::p_load(tidyverse, ggpubr)
set.seed(123)
n <- 50
data <- data.frame(x = sample(c("NY", "PA", "NJ"), n, replace = T), y = sample(c("Male", "Female"), n, replace = T), t = sample(0:1, n, replace = T))
data %>% ggplot(aes(x = as.factor(x), y = as.factor(y), color = as.factor(t), shape = as.factor(t))) +
  geom_point(position = position_jitter(width = 0.1, height = 0.1), size=2) +
  labs(
    x = "",
    y = "",
    color = "treat",
    shape = "treat"
  ) +
  scale_color_manual(values = c("0" = "#D55E00", "1" = "#0072B2")) +
  scale_shape_manual(values = c("0" = 16, "1" = 17)) +
  theme_minimal()
```

# Coarsened Exact Matching

- Coarsen continuous covariates into discrete bins
- Then exact match on the bin values.
\pause

- Problems:
  - Sensitive to bin size
  - Still might be bins with no control units
  - Overlap will determine how narrow the bins can be
- Ideal when you have lots of data and relatively few covariates


# Coarsened Exact Matching, illustration

```{r}
#| echo: false
#| warning: false
#| fig-width: 8
#| fig-height: 4
 
set.seed(123)
n <- 200
data <- data.frame(x = rnorm(n, 30, sd = 5), y = runif(n, 0, 12), t = sample(0:1, n, replace = T))
plt <- data %>% ggplot(aes(x = x, y = y, color = as.factor(t), shape = as.factor(t))) +
  geom_point(size=2) +
  labs(
    x = "Age",
    y = "Education",
    color = "treat",
    shape = "treat"
  ) +
  scale_color_manual(values = c("0" = "#D55E00", "1" = "#0072B2")) +
  scale_shape_manual(values = c("0" = 16, "1" = 17)) +
  theme_minimal() +
  theme(panel.grid.minor = element_blank(), panel.grid.major = element_blank())
ggarrange(plotlist = list(
  plt,
  plt + theme(panel.grid.major = element_line(
    color = "#8ccde3",
    size = 0.75,
    linetype = 2
  )),
  plt + scale_x_continuous(breaks = seq(20, 45, 8)) + scale_y_continuous(breaks = seq(0, 12, 3)) + theme(panel.grid.major = element_line(
    color = "#8ccde3",
    size = 0.75,
    linetype = 2
  ))
), nrow = 1, common.legend = T, legend = "bottom")
```

# Nearest Neighbor Matching

- Select close units by minimizing some function of the covariates 
- Generally create a matched sample of units, and then estimate ATT/ATE 
- Researchers can try different matching algorithms until an acceptable level of balance is achieved
  - **Balance Checks**
\pause

- Formally for **ATT**:
$$\hat{\tau}_m = \frac{1}{n_1} \sum_{i=1}^{n} D_i \left( Y_i - \frac{1}{M} \sum_{j \in \mathcal{J}_m(i)} Y_j \right)$$
- $\mathcal{J}(i)$ are the set of $M$ closest control units to $i$ in terms of $X_i$.
- $n_1$ number of treated units

# Nearest Neighbor Matching
First, Define closeness in covariate space using scalar distance metric \pause

**Mahalanobis distance**: 
generalization of Euclidean distance that handles correlated features and different variances
  $$d_M(X_i, X_j)=\sqrt{(X_i - X_j)^{\prime} S^{-1} (X_i - X_j)},$$
where $S$ is a covariance matrix
\pause

**Propensity score**: 
$$e(X_i) := \Pr\!\left[D_i = 1 \mid X_i\right] = \mathbb{E}\!\left[ D_i \mid X_i \right].$$

  - Under strong ignorability, $D_i \perp Y_{di} \mid e(X_i)$.
  - Matching on $e(X_i)$ is sufficient for identification. \pause

Then, match using the "nearest neighbors" by this distance metric

# NN matching choices

- How many control units per treated?
  - Lower reduces bias
  - More reduces variance
\pause

- With or without replacement: same control matched to multiple treated?
  - **With replacement** minimizes the covariate distance between matched units

# NN matching bias correction
- NN matchingg estimator has generally has asymptotic bias.

- Suppose a treated unit $i$ and matched control $j$.
- Matching estimate of the treatment effect $\tau$:
\begin{align*}
\hat{\tau} & = Y_{i1} - Y_{j0} \\ 
& = Y_{i1} - Y_{i0} + Y_{i0} - Y_{j0}\\
& = \tau + (Y_{i0} - Y_{j0}).
\end{align*} \pause
- Estimate $\hat{Y}_{i0}, \hat{Y}_{j0}$ (e.g., with OLS but can also use ML methods). \pause

**Bias corrected**
$$\tilde{\tau} = \hat{\tau} - (\hat{Y_{i0}} - \hat{Y_{j0}}) \quad \text{(Abadie and Imbens, 2011)}.$$

# Evaluating balance

- Differences-in-means/medians, standardized.
- QQ plots/KS statistics for comparing the entire distribution of $X_i$.
- $L_1$: multivariate histogram (for CEM)
- Choice of metric can change what matching method works best.

\pause

Test for balance

- Do not use simple null hypothesis to test for balance 
  - Dropping units can lower power (increase p-values) without a change in balance.
- To test for balance, use an "equivalence test" to see if you could reject some degree of imbalance ( Hartman and Hidalgo 2017)

\href{https://cran.r-project.org/web/packages/MatchIt/vignettes/assessing-balance.html}{`MatchIt` Assessing Balance Page}

# Matching packages

- `MatchIt` performs exact matching, genetic, CEM, and NN
  - Different distance metrics allowed
- Other packages: `Matching`, `GenMatch`, `cem`

# NN Matching example
\href{https://www.jstor.org/stable/44651934}{Ruggeri, Dorussen, and Gizelis (2016)}
```{r}
#| message: false
#| warning: false
#| echo: true

# Import the data
library(haven)
data <- read_dta("matchingdata.dta")

# Keep non-missing values
data <- na.omit(data)

# Treatment distribution
library(janitor)
tabyl(data$PKO)
```
```{r setup}
# scipen = 999 turns off scientific notation
options(scipen = 999)
# Optionally set digits globally for printing
options(digits = 2)
```

# NN Matching example
\scriptsize
```{r}
#| message: false
#| warning: false
#| echo: true

# Use MatchIt package
library(MatchIt)

# Initial imbalance
nn_raw <- matchit(PKO ~ avgttime + avgadjimr + popgpw2000_40 + 
                      avgmnt +  borddist + capdist + prec_new,
                  data = data,
                  method = NULL)

knitr::kable(summary(nn_raw)[[3]][, 1:4, drop = FALSE])
```


# NN Matching example
\scriptsize
```{r}
#| message: false
#| warning: false
#| echo: true

# Nearest-neighbor matching on the propensity score
nn_match <- matchit(PKO ~ avgttime + avgadjimr + popgpw2000_40 + 
                      avgmnt +  borddist + capdist + prec_new, 
                    method = "nearest", 
                    data = data)

nn_match
# Matched sample
data_nn <- match.data(nn_match)
# Matched data
tabyl(data_nn$PKO)
```

# NN Matching example
\scriptsize
```{r}
#| message: false
#| warning: false
#| echo: true
summary(nn_match)$sum.all[,1:3]
summary(nn_match)$sum.matched[,1:3]
```

# NN Matching example
\scriptsize
```{r}
#| message: false
#| warning: false
#| echo: true
#| fig-width: 8
#| fig-height: 4

plot(nn_match, type = "jitter", interactive = FALSE)
```

# NN Matching example
\scriptsize
```{r}
#| message: false
#| warning: false
#| echo: true
nn_match_wr <- matchit(PKO ~ avgttime + avgadjimr + popgpw2000_40 + 
                      avgmnt +  borddist + capdist + prec_new, 
                    method = "nearest", 
                    replace = TRUE,
                    data = data)
nn_match_wr

# Matched sample
data_nn_wr <- match.data(nn_match_wr)
# Matched data
tabyl(data_nn_wr$PKO)
```

# NN Matching example
\scriptsize
```{r}
#| message: false
#| warning: false
#| echo: true
summary(nn_match)$sum.matched[,1:3]
summary(nn_match_wr)$sum.matched[,1:3]
```

# NN Matching example
\scriptsize
```{r}
#| message: false
#| warning: false
#| echo: true
#| fig-width: 8
#| fig-height: 4
plot(nn_match_wr, type = "jitter", interactive = FALSE)
```

# NN Matching example
\scriptsize
```{r}
#| message: false
#| warning: false
#| echo: true
#| fig-width: 8
#| fig-height: 4
plot(nn_match_wr, type = "density", interactive = FALSE, 
     which.xs = ~avgttime + avgadjimr)
```


# NN Matching example
\scriptsize
```{r}
#| message: false
#| warning: false
#| echo: true
#| fig-width: 8
#| fig-height: 4

library(ggplot2)
# Raw 
ggplot(data, aes(x=avgttime, y = ..density.., fill=PKO)) + geom_histogram(bins=50) +
  facet_wrap(~PKO)+ theme_bw() +
  theme(legend.position="none")
```
# NN Matching example
\scriptsize
```{r}
#| message: false
#| warning: false
#| echo: true
#| fig-width: 8
#| fig-height: 4

# matched 
ggplot(data_nn_wr, aes(x=avgttime, y = ..density.., fill=PKO)) + geom_histogram(bins=50) +
  facet_wrap(~PKO)+ theme_bw() +
  theme(legend.position="none")
```


# DAGs in R `dagitty`
```{r}
#| echo: true
library(dagitty)

g1 <- dagitty( "dag {
    U -> X -> Y
    X -> D -> Y <- V -> X
}")

children(g1, "X")

parents(g1, "D")
```


# `dagitty`
```{r}
#| echo: true
#| fig-align: center
#| fig-width: 8
#| fig-height: 4

plot(graphLayout(g1))
```

# `dagitty`
```{r}
#| echo: true

print( impliedConditionalIndependencies( g1 ) )
```


# `dagitty`

```{r}
#| echo: true

print( adjustmentSets( g1, "D", "Y", effect="direct" ) )
```


# `dagitty`

\tiny
```{r}
#| echo: true

for( n in names(g1) ){
    for( m in children(g1,n) ){
        a <- adjustmentSets( g1, n, m, effect="direct" )
        if( length(a) > 0 ){
            cat("The coefficient on ",n,"->",m,
                " is identifiable controlling for:\n",sep="")
            print( a, prefix=" * " )
        }
    }
}
```

# `dagitty`
```{r}
#| echo: true

g2 <- dagitty( "dag {
    B -> T -> F -> Y
    T -> E -> Y
    B -> D -> E -> Y <- D
    B -> C -> Y
    B -> C <- A -> Y
}")
```


# `dagitty`
```{r}
#| echo: true
#| fig-align: center
#| fig-width: 8
#| fig-height: 4

plot(graphLayout(g2))

```


# `dagitty`
```{r}
#| echo: true

print(adjustmentSets(g2, "T", "Y", effect="total" ) )
```


# `dagitty`
```{r}
#| echo: true
g3 <- dagitty( "dag {
    Z -> D -> Y <- X -> D
    U -> D -> Y <- U
}")

```

# `dagitty`
```{r}
#| echo: true
#| fig-align: center
#| fig-width: 8
#| fig-height: 4

plot(graphLayout(g3))

```


# DAGs in R
```{r}
#| echo: true

print(instrumentalVariables(g3, "D", "Y"))
```