---
title: "Quant 2, Week 8 Lab"
subtitle: "Panel Data and Difference in Differences"
author: "Sylvan Zheng"
output: pdf_document
---

# TWFE Staggered event bias

- TWFE bias with multiple events
- Go over cyrus' code
- Introduce packages for staggered treatment and sensitivity analysis


```{r}
pacman::p_load(tidyverse, knitr, fixest)
pacman::p_load(did, HonestDiD, haven, panelView)

set.seed(123)
periods <- 5
N <- 100

# Generate simulation data.
# D.pat is a vector of length period of the treatment timing
# YC.pat is a vector of length period of the mean Y for each unit
# YT.func is a function that returns a vector of length period
#    of the exact change in Y for each unit
#    This is a function so that you can include calls to rnorm
generate_data <- function(N, first.treat, YC.pat, YT.func) {
    # Fancy code that basically just returns a vector like c(0,0,1,1,1) if first.treat = 3
    D.pat <- c(rep(0, first.treat - 1), rep(1, periods - first.treat + 1))[1:periods]

    tibble(
        id = first.treat * N + rep(1:N, each = periods),
        period = rep(1:periods, N),
        D = rep(D.pat, N),
        YC = rep(YC.pat, N) + rnorm(N * periods),
        YT = YC + (replicate(N, YT.func()) %>% as.vector())
    ) %>%
        mutate(
            Y = if_else(D == 1, YT, YC),
            first.treat = first.treat
        )
}

# Never-treated group.
dat_U <- generate_data(
    N,
    first.treat = 6,
    YC.pat = c(0, 0, 1, 2, 3),
    YT.func = \() c(0, 0, 0, 0, 0)
)
dat_2 <- generate_data(
    N,
    first.treat = 2,
    YC.pat = 2 + c(0, 0, 1, 2, 3),
    YT.func = \() c(0, 1, rnorm(1) + 2, rnorm(1) + 4, rnorm(1) + 8)
)
dat_5 <- generate_data(
    N,
    first.treat = 5,
    YC.pat = 3 + c(0, 0, 1, 2, 3),
    YT.func = \() c(0, 0, 0, 0, rnorm(1) + 1)
)

dat_all <- rbind(dat_U, dat_2, dat_5) %>% mutate(group = as.factor(first.treat))

# Plot the data

# First, the YCs
dat_all %>% ggplot(aes(x = period, y = YC, color = group)) +
    geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
    stat_summary(fun = mean, geom = "line")

# Then, the Ys
dat_all %>% ggplot(aes(x = period, y = Y, color = group)) +
    geom_point(position = position_jitter(width = 0.1), alpha = 0.6) +
    stat_summary(fun = mean, geom = "line")
```

## True ATTs

```{r}
dat_all %>%
    filter(D == 1) %>%
    group_by(group, period) %>%
    summarize(att = mean(YT) - mean(YC)) %>%
    kable(digits = 2)

dat_all %>%
    group_by(group, period) %>%
    summarize(att = mean(YT) - mean(YC)) %>%
    ggplot(aes(x = period, y = att, color = group, shape = group)) +
    geom_line() +
    geom_point()

# Average ATT
dat_all %>%
    filter(D == 1) %>%
    group_by(group, period) %>%
    summarize(att = mean(YT) - mean(YC)) %>%
    .$att %>%
    mean()
```

## TWFE

```{r}
twfe <- feols(Y ~ D | id + period, data = dat_all)
etable(twfe)
```

## Event by Event

```{r}
event1.2 <- dat_all %>% filter(
    group %in% c(2, 6),
    period %in% c(1, 2)
)
event1.3 <- dat_all %>% filter(
    group %in% c(2, 6),
    period %in% c(1, 3)
)
event1.4 <- dat_all %>% filter(
    group %in% c(2, 6),
    period %in% c(1, 4)
)
event1.5 <- dat_all %>% filter(
    group %in% c(2, 6),
    period %in% c(1, 5)
)
event2.5 <- dat_all %>% filter(
    group %in% c(5, 6),
    period %in% c(1, 5)
)

models <- list(
    feols(Y ~ D | id + period, data = event1.2),
    feols(Y ~ D | id + period, data = event1.3),
    feols(Y ~ D | id + period, data = event1.4),
    feols(Y ~ D | id + period, data = event1.5),
    feols(Y ~ D | id + period, data = event2.5)
)
etable(models)
# Looks familiar? Compare to the true ATTs

# Each event has an equal number of observations, so the naive mean is fine
mean(sapply(models, \(m) coef(m)["D"]))
```

# Tools for estimating with staggered event

## Package fixest: Sun and Abraham (2021)

```{r}
sa.model <- feols(Y ~ 1 + sunab(first.treat, period) | period + id, data = dat_all, cluster = "id")
etable(sa.model)
# Plot the event study
iplot(sa.model)
# Get the ATT
summary(sa.model, agg = "ATT")
# SE 0.13
summary(sa.model, agg = "cohort")
```

## Package did: Callaway and Santa Anna (2021)

```{r}
out <- att_gt(
    yname = "Y",
    gname = "first.treat",
    idname = "id",
    tname = "period",
    xformla = ~1, # if you had covariates for conditional parallel trends, put here
    data = dat_all,
    est_method = "reg"
)
summary(out)
ggdid(out)
ggdid(aggte(out, type = "dynamic"))
summary(aggte(out, type = "simple"))
summary(aggte(out, type = "group"))
```

# Application

```{r}
df <- read_dta("https://raw.githubusercontent.com/Mixtape-Sessions/Advanced-DID/main/Exercises/Data/ehec_data.dta")
data(mpdta)

head(mpdta)
# county-year panel
# lemp: log employment
# lpop: log population
# first.treat: year of minimum wage increase
mpdta$first.treat %>% table(useNA = "ifany")
mpdta$year %>% table(useNA = "ifany")

## Panelview
mpdta <- mpdta %>% mutate(
    D = case_when(
        first.treat == 0 ~ 0,
        TRUE ~ first.treat <= year
    ),
    ftsa = case_when(
        first.treat == 0 ~ 9999,
        T ~ first.treat
    ),
    treat = first.treat != 0
)
panelview(mpdta, Y = "lemp", D = "D", index = c("countyreal", "year"))
```

## Estimate effects

```{r}
# Estimate effects for the following specifications:

# 1. TWFE comparing the 2006 cohort to the never-treated cohort
# 2. Sun Ab
# 3. Callaway and Sant'anna
# 4. Callaway and Sant'anna incorporating lpop as a covariate

```

## Incorporate Covariates

```{r}
# Sun Ab
sa.model <- feols(lemp ~ sunab(ftsa, year) | countyreal + year, data = mpdta)
ggiplot(sa.model)
# CS
cs.model <- att_gt(
    yname = "lemp",
    gname = "first.treat",
    idname = "countyreal",
    tname = "year",
    xformla = ~lpop,
    data = mpdta,
    est_method = "reg"
)
ggdid(aggte(cs.model, type = "dynamic"))
```

## Sensitivity analysis: HonestDiD

### TWFE

```{r}
betahat <- summary(twfe)$coefficients # save the coefficients
sigma <- summary(twfe)$cov.scaled # save the covariance matrix
delta_rm_results <-
    createSensitivityResults_relativeMagnitudes(
        betahat = betahat, # coefficients
        sigma = sigma, # covariance matrix
        numPrePeriods = 3, # num. of pre-treatment coefs
        numPostPeriods = 1, # num. of post-treatment coefs
        Mbarvec = seq(0.5, 2, by = 0.5) # values of Mbar
    )
delta_rm_results
originalResults <- constructOriginalCS(
    betahat = betahat,
    sigma = sigma,
    numPrePeriods = 3,
    numPostPeriods = 1
)
createSensitivityPlot_relativeMagnitudes(delta_rm_results, originalResults)
```

### Staggered adoption and honest DiD

```{r}
# Load these functions because HonestDiD and SunAb/CallawaySantanna are not well integrated
source(here::here("lab08/honest-helpers.R"))
```

#### SunAb

```{r}
svcv <- sunab_beta_vcv(sa.model)
betahat <- svcv$beta
sigma <- svcv$sigma
delta_rm_results <-
    createSensitivityResults_relativeMagnitudes(
        betahat = betahat,
        sigma = sigma,
        numPrePeriods = 3,
        numPostPeriods = 4,
        Mbarvec = seq(0.5, 2, by = 0.5)
    )
originalResults <- constructOriginalCS(
    betahat = betahat,
    sigma = sigma,
    numPrePeriods = 3,
    numPostPeriods = 4
)
createSensitivityPlot_relativeMagnitudes(delta_rm_results, originalResults)

# Honest DiD default does this for the first period. But what if we are interested in all the effects?

delta_rm_results <-
    createSensitivityResults_relativeMagnitudes(
        betahat = betahat,
        sigma = sigma,
        numPrePeriods = 3,
        numPostPeriods = 4,
        Mbarvec = seq(0.5, 2, by = 0.5),
        l_vec = c(0.25, 0.25, 0.25, 0.25)
    )
originalResults <- constructOriginalCS(
    betahat = betahat,
    sigma = sigma,
    numPrePeriods = 3,
    numPostPeriods = 4,
    l_vec = c(0.25, 0.25, 0.25, 0.25)
)
createSensitivityPlot_relativeMagnitudes(delta_rm_results, originalResults)
```

### Callaway Sant'anna

```{r}
# Need to specify the "universal" type base period to use honestDiD
cs.model <- att_gt(
    yname = "lemp",
    gname = "first.treat",
    idname = "countyreal",
    tname = "year",
    xformla = ~lpop,
    data = mpdta,
    est_method = "reg",
    base_period = "universal"
)
ggdid(aggte(cs.model, type = "dynamic"))

# The helper code automatically does the entire post periods. No l_vec here
sensitivity_results <-
    honest_did(aggte(cs.model, type = "dynamic"),
        e = 0,
        type = "relative_magnitude",
        Mbarvec = seq(from = 0.5, to = 2, by = 0.5)
    )
createSensitivityPlot_relativeMagnitudes(
    sensitivity_results$robust_ci,
    sensitivity_results$orig_ci
)
```
